---
title: "Explainable Boosting Machine for MX Lead Conversion"
author: "Josh McAlister"
date: "2026-04-17"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"
    embed-resources: true
execute:
  include: true
  eval: true
  warning: false
  message: false
editor:
  markdown:
    wrap: sentence
---

# About This Notebook

This notebook contains my individual contribution to the MasterControl MX Lead Progression group capstone project: an Explainable Boosting Machine (EBM) model that predicts whether a qualified Mx lead will progress to SQL, SQO, or Won.
The full group project, including the shared baseline GLM and decision tree pipeline that this model was designed to fit alongside, lives at <https://github.com/gabyr93/Group-Project-Master-Control>.

The EBM was designed to be fully self-contained.
It reads the modeling-ready dataset produced in the group's EDA phase, joins the cluster-label features I engineered from the clustering analyses, and creates its own train/test split using the same `set.seed(123)` + `caret::createDataPartition()` call as the shared pipeline — guaranteeing identical row assignments so AUC comparisons across the three models are fair.

> **Note on reproducibility:** The project's source data (`Mastercontrol QAL Performance.csv`) is not committed to GitHub per data-privacy requirements.
> The `.rds` files referenced below are produced by the EDA notebook in the group repo.
> To re-render this notebook locally, place `leads_cleaned_final.rds`, `cluster_labels.rds`, and `mx_cluster_labels.rds` in a `data/` subfolder.

# Explainable Boosting Machine (EBM)

In this section, we fit an Explainable Boosting Machine (EBM) as the third modeling approach in the project.
EBMs are a type of Generalized Additive Model with pairwise interactions (GA²M) that use cyclic gradient boosting to learn shape functions for each feature.
Unlike black-box models, EBMs are fully interpretable — each feature's contribution can be visualized independently — while still achieving accuracy competitive with ensemble methods like random forests and gradient boosting.

EBMs are especially useful for this project because they can show exactly *how* each lead characteristic (industry, territory, title role, enrichment quality) relates to conversion probability, which directly supports the business goal of understanding what makes a strong Mx lead.

This section uses the `ebm` R package, which wraps the Python `interpret` library via `reticulate`.

```{r setup-libs, include=FALSE}
library(tidyverse)
library(caret)
```

```{r EBM Setup, include=FALSE}
# Wire up the reticulate/Python backend the ebm package depends on.
# The virtualenv only needs to be created once per machine — the install
# call is a no-op if the packages are already present.
library(reticulate)
use_virtualenv("r-ebm", required = TRUE)
virtualenv_create("r-ebm")
virtualenv_install("r-ebm", packages = c("interpret", "numpy", "scikit-learn"))
library(ebm)
library(patchwork)
```

```{r EBM Data Prep}
# Load cluster labels exported from the EDA phase
cluster_labels    <- readRDS("data/cluster_labels.rds")
mx_cluster_labels <- readRDS("data/mx_cluster_labels.rds")

# Build the EBM dataset independently — join cluster features, filter to Mx
ebm_data <- readRDS("data/leads_cleaned_final.rds") %>%
  left_join(cluster_labels, by = "qal_id") %>%
  left_join(mx_cluster_labels, by = "qal_id") %>%
  filter(solution_rollup == "Mx") %>%
  mutate(
    next_stage_target = factor(
      next_stage_target,
      levels = c("0", "1"),
      labels = c("Not Successful", "Successful")
    )
  ) %>%
  # Replace NAs in title_cluster with an explicit factor level.
  # Rows flagged as title_lowinfo were never clustered, so they carry NA.
  # EBM handles NAs internally but cannot display them in shape-function plots.
  mutate(
    title_cluster = fct_na_value_to_level(title_cluster, level = "No_Title_Info")
  ) %>%
  select(-qal_id, -contact_lead_id, -next_stage_c,
         -contact_lead_title, -solution_rollup)

# Drop a leftover duplicate column if the source data contained one
if ("solution.1" %in% names(ebm_data)) {
  ebm_data <- ebm_data %>% select(-solution.1)
}

# Same seed + stratified split as the shared pipeline -> identical row partition
set.seed(123)
ebm_split_idx <- caret::createDataPartition(
  y = ebm_data$next_stage_target, p = 0.70, list = FALSE
)
ebm_train <- ebm_data[ebm_split_idx, ]
ebm_test  <- ebm_data[-ebm_split_idx, ]

dim(ebm_train)
dim(ebm_test)
```

### Interpretation of EBM Data Prep

The EBM dataset is built independently from the shared pipeline to avoid introducing cluster-label NAs into the group's existing GLM and decision tree recipes.
Three engineered features from the EDA clustering phase — `account_cluster`, `title_cluster`, and `mx_cluster` — are joined onto the base modeling data.
For `title_cluster`, rows where the lead title was flagged as low-info (and therefore never clustered) have their NA replaced with an explicit `"No_Title_Info"` factor level so that EBM can visualize the shape function for that group.

The train/test split uses the same `set.seed(123)` and `caret::createDataPartition()` call as the shared pipeline, which guarantees identical row selection since the function depends only on the target vector and the seed.
This means the EBM is evaluated on exactly the same holdout set as the GLM and decision tree, making AUC comparisons fair.

```{r EBM Fit}
# Fit the EBM. The formula interface is handled by the ebm package; the
# model auto-detects pairwise interactions and fits shape functions for
# each main effect.
mx_ebm <- ebm(next_stage_target ~ ., data = ebm_train, objective = "log_loss")
print(mx_ebm)
```

### Interpretation of EBM Fit

The `ebm()` call fits a GA²M model using cyclic gradient boosting.
Each feature gets its own shape function (main effect), and the model also automatically detects and fits the most important pairwise interaction terms.
The `log_loss` objective is appropriate for binary classification — it optimizes the model to produce well-calibrated predicted probabilities rather than just class labels.

The printed summary shows the number of main effects and interaction terms retained, along with the intercept.
Unlike the GLM (which required dummy encoding via `step_dummy`) or the decision tree (which splits on one variable at a time), the EBM learns flexible nonlinear relationships for each feature while remaining fully interpretable.

```{r EBM Predictions}
# Generate predictions on the held-out test set
ebm_probs  <- predict(mx_ebm, newdata = ebm_test)
ebm_labels <- predict(mx_ebm, newdata = ebm_test, type = "class")

# Confusion matrix — show how the model classifies each test lead
table("Observed" = ebm_test$next_stage_target, "Predicted" = ebm_labels)

# ROC AUC via yardstick so it matches the metric reported by the
# GLM and decision tree in the shared pipeline (val.prob below
# reports the same number as C-statistic — they are mathematically
# identical for binary classification).
ebm_roc <- tibble(
  truth    = ebm_test$next_stage_target,
  estimate = ebm_probs[, 2L]
)
yardstick::roc_auc(ebm_roc, truth, estimate, event_level = "second")
```

### Interpretation of EBM Predictions

The confusion matrix shows how the EBM classifies test-set leads compared to their actual outcomes.
The key metrics to watch are:

-   **True Positives (Successful predicted as Successful):** how many real successes the model catches
-   **False Negatives (Successful predicted as Not Successful):** how many real successes the model misses
-   **False Positives (Not Successful predicted as Successful):** leads the model incorrectly flags as promising

Because the target is imbalanced (\~12.9% success rate), raw accuracy can be misleading — a model that predicts "Not Successful" for every lead would still be \~87% accurate.
The calibration plot in the next section provides a better view of whether the predicted probabilities are reliable.

```{r EBM Calibration}
library(rms)
# Calibration plot: predicted probability vs. observed frequency.
# This reports the C-statistic (ROC AUC) and Brier score, plus a visual
# check that predicted probabilities align with actual outcome rates.
par(las = 1)
rms::val.prob(
  ebm_probs[, 2L],
  as.integer(ebm_test$next_stage_target == "Successful"),
  cex = 0.5
)
```

### Interpretation of EBM Calibration

The calibration plot compares the model's predicted probabilities against the actual observed success rates.
A well-calibrated model should have points falling close to the 45-degree diagonal line — meaning that when the model says a lead has a 20% chance of success, roughly 20% of those leads actually succeed.

Key statistics reported on the plot include the **C-statistic (0.869)**, which is equivalent to ROC AUC and directly comparable to the group's GLM (0.79) and decision tree (0.82).
The **Brier score (0.073)** is low, indicating well-calibrated predicted probabilities.
The **calibration slope (0.98)** is near 1.0, confirming that predicted probabilities correspond tightly to observed frequencies.

```{r EBM Global Explanation}
# Top 15 most important terms by mean absolute score (main effects + interactions)
plot(mx_ebm, n_terms = 15)

# Shape function for a single feature — shows how each category of
# acct_territory_rollup shifts the prediction up or down
plot(mx_ebm, term = "acct_territory_rollup", horizontal = TRUE)

# List the pairwise interactions the model detected
# (capped at 20 rows so the document compiles cleanly)
head(mx_ebm$term_names_[grepl("&", mx_ebm$term_names_)], n = 20)
```

### Interpretation of EBM Global Explanation

The term importance plot ranks each feature (and interaction term) by its mean absolute contribution to the predicted log-odds of success.
Features at the top of the chart have the largest overall influence on whether a lead is predicted as successful or not.
In this model, `priority` (engagement type) is the single strongest predictor by a wide margin, followed by the data-completeness flags `mfg_lowinfo` and `title_lowinfo`, then the engineered `account_cluster` and `title_cluster` features.

The shape function plot for `acct_territory_rollup` shows exactly how each territory category shifts the prediction — Americas is the only territory with a positive score, while every other region pulls the prediction downward.
Unlike a logistic regression coefficient (which gives a single number), the EBM shape function can show nonlinear and non-monotonic relationships across categorical levels.

The listed pairwise interactions reveal which feature combinations the model found to be jointly predictive beyond their individual effects.
These interactions can surface insights that single-variable analyses miss — for instance, a specific industry × territory combination that converts unusually well.

```{r EBM Local Explanation}
# Local explanation for a single test-set lead. The waterfall plot
# decomposes the prediction into each feature's contribution, starting
# from the intercept and adding/subtracting shape-function values.
newx <- subset(ebm_test[1L, ], select = -next_stage_target)
newy <- ebm_test$next_stage_target[1L]
plot(mx_ebm, local = TRUE, X = newx, y = newy)
```

### Interpretation of EBM Local Explanation

The local explanation waterfall plot decomposes the prediction for a single lead into the contributions from each feature.
Starting from the intercept (the baseline log-odds of success), each bar shows how much a specific feature pushes the prediction up (toward Successful) or down (toward Not Successful).

This is especially useful for the business team because it can answer questions like: *"Why did the model score this particular lead highly?"* or *"What would need to change about this lead for it to be predicted as successful?"* Each local explanation is fully faithful to the model — it is not an approximation — because EBMs are additive by design.
That exact-arithmetic property is the core reason the EBM was selected over XGBoost: both achieved strong ranking performance, but only the EBM can tell a sales rep *why* a given lead received its score.

```{r EBM Save}
# Save the fitted model as a Python pickle. EBM objects are reticulate
# wrappers around Python classes, so saveRDS does not work across the
# R/Python bridge — py_save_object is the supported path.
reticulate::py_save_object(mx_ebm, filename = "data/mx_ebm.pkl")
```

# Summary

The EBM achieved a test-set ROC AUC of **0.87**, outperforming the group's XGBoost (0.82), decision tree (0.76), and logistic regression (0.79) models while remaining fully interpretable.
The calibration slope of 0.98 and Brier score of 0.073 confirm the predicted probabilities are trustworthy for business thresholding decisions.

The three most important signals in the model are:

1.  **Priority / engagement type** — how a lead engaged (Contact Us, Video Demo, Live Demo vs. Webinar Demo, Priority 2)
2.  **Data completeness** — missing enrichment fields (`mfg_lowinfo`, `title_lowinfo`) are themselves strong negative signals
3.  **Job title cluster** — role-based segments derived from unsupervised clustering of lead titles

For the full business narrative, ideal customer profile, hypothetical scoring scenarios, and cost-benefit analysis, see the group's repository: <https://github.com/gabyr93/Group-Project-Master-Control>.
